Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-17 paper/ms63.tex
a few more edits

\racomment{
In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
gyrochronology model is plotted for comparison with the data.
This model was chosen as an example but there are many other empirical
gyrochronology models we could have used.
All the available empirical gyrochronology models are similar in essence to
the \citet{angus2019} relation and, like that relation, none is able to
reproduce the observed velocity dispersions, or capture the evolving shape of
the period-\teff\ relations.
The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.

Alternatively, we could have chosen to compare our data with one of the many
other available empirical gyrochronology models.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ (around
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data.
It has a similar period-\teff\ relation to the Praesepe-based
\citet{angus2019} model, because the period-\teff\ relation of Praesepe
roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010}
gyrochrones do not reproduce the data explored here and do not have the
flexibility to capture mass and time-variable rotational evolution.

The lack of flexibility that is baked-in to these empirical gyrochronology
models should not be seen as a failing of gyrochronologists: with a paucity of
calibration data for old stars all one can do is fit simple relations to young
stars and extrapolate to old ages.
Naturally however, in this era of precise space-based photometry, new data are
revealing flaws in our old models (which no one ever thought were perfect),
and are providing opportunities to improve them.
}

@@

\racomment{
In order to measure a rotation period at all, there must be some magnetically
active regions (either bright plages or dark spots), of a reasonable size on
the surface of a star.
    The relatively sharp upper-edge of the rotation period distribution seen
in figure \ref{fig:vplot} may therefore be caused by a finite active lifetime
of stars.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
kiman2019}, and if the detectability of a rotation period is considered to be
a magnetic activity proxy, then our results provide further evidence for a
{\it mass-dependent} activity lifetime.
Figure \ref{fig:vplot} shows that the populations of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
}
% a decrease in magnetic activity after stars exceed a certain age.
% If so, it is possible that the ages of stars near this upper-edge represent
% the activity lifetimes of stars.}
% % We therefore assume that stars {\it with} measured rotation periods are,
% on % average, more magnetically active than stars {\it without} measured
% rotation % periods.
% \citet{matt2015} found that a 4 Gyr gyrochrone, calculated with their model,
% coincided with the upper-envelope at 4 Gyrs and concluded that the
% upper-envelope of the \kepler\ rotation period distribution corresponds to a
% single-age population of around 4 Gyr.
% However, the velocity dispersion data presented in figure \ref{fig:vplot}
% suggest that the activity lifetime of stars is mass-dependent.
% In this case, stars near the upper-dge of the period distribution are not all
% the same age that later-type stars are in fact older than earlier types.
% }

>>>>

\racomment{
In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
gyrochronology model is plotted for comparison with the data.
This model was chosen as an example but there are many other empirical
gyrochronology models we could have used.
All the available empirical gyrochronology models are similar in essence to
the \citet{angus2019} relation and, like that relation, none is able to
reproduce the observed velocity dispersions, or capture the evolving shape of
the period-\teff\ relations.

The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.
There are many other empirical gyrochronology models we could have chosen to
compare with our data.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data here.
The \citet{barnes2010} model has a similar period-\teff\ relation to the
Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
Praesepe roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
in its current form does not have the flexibility to capture the mass and
time-variable rotational evolution revealed by the data explored here.

The lack of flexibility in these empirical gyrochronology models should not be
seen as a failing of gyrochronologists.
With a lack of old calibration stars, the only possible approach has been to
fit simple relations to young stars and extrapolate to old ages.
Naturally however, in this era of precise space-based photometry, new data are
revealing flaws in our old models (which no one ever thought were perfect),
and are providing opportunities to improve them.
}

@@

\racomment{
In order to measure a rotation period at all, there must be some magnetically
active regions (either bright plages or dark spots), of a reasonable size on
the surface of a star.
In other words, there is presumably some minimum magnetic activity level that
is required in order to produce coherent light curve variability from which a
rotation period can be confidently measured.
The relatively sharp upper-edge of the rotation period distribution seen in
figure \ref{fig:vplot} may be caused by a finite active lifetime of stars,
\ie\ stars older than a critical age are no longer active enough to produce
periodic, high-amplitude variability.
The data presented in figure \ref{fig:vplot} indicate that this active
lifetime is a function of stellar mass.
The mass dependence of magnetic activity lifetimes has been demonstrated
previously \racomment{for M dwarfs} \citep[\eg][]{west2008, newton2017,
kiman2019}, and if the detectability of a rotation period is considered to be
a magnetic activity proxy, then our results provide further evidence for a
{\it mass-dependent} activity lifetime.
Figure \ref{fig:vplot} shows that the populations of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that \racomment{most of} the oldest stars with detectable
rotation periods are cooler than 4500 K, \ie\ these-low mass stars stay active
longer than more massive stars.
}
% a decrease in magnetic activity after stars exceed a certain age.
% If so, it is possible that the ages of stars near this upper-edge represent
% the activity lifetimes of stars.}
% % We therefore assume that stars {\it with} measured rotation periods are,
% on % average, more magnetically active than stars {\it without} measured
% rotation % periods.
% \citet{matt2015} found that a 4 Gyr gyrochrone, calculated with their model,
% coincided with the upper-envelope at 4 Gyrs and concluded that the
% upper-envelope of the \kepler\ rotation period distribution corresponds to a
% single-age population of around 4 Gyr.
% However, the velocity dispersion data presented in figure \ref{fig:vplot}
% suggest that the activity lifetime of stars is mass-dependent.
% In this case, stars near the upper-dge of the period distribution are not all
% the same age that later-type stars are in fact older than earlier types.
% }
